Procedural Terrain \\ {\large Final project for class CS6491 Computer Graphics}

### 2015-11-26 Paper/Paper.tex
implemented smoothing limit
finished diffusion / terrain feature chapter

Each feature curve consists of two or more control points. Each control point has the following properties:
\begin{itemize}
	\item a position $(x,y,z)$ in the world
	\item a boolean if it constraints elevation
	\item a float $plateau$, specifying the size of the flat plateau on top of the curve
	\item four floats $s_l, \alpha_l, s_r, \alpha_r$ specifying the size and angle of the slopes left and right of the curve
\end{itemize}
In the example in Fig. \ref{fig:diffusion1}, the slope sizes are zero, so no slope constraints are applied.

@@

\subsection{Diffusion solver}\label{DiffusionSolver}
The task of the diffusion solver is to modify the terrain so that it remains smooth while preserving the feature constraints.
The diffusion solver iteratively updates the terrain using the following recursion: Let $H^i$ be the height map at the i-th iteration.
\begin{equation}
	H^{k+1}[i,j] = \alpha E[i,j] + \beta G^{k+1}[i,j] + (1-\alpha-\beta) L^{k+1}[i,j]
\label{eq:Diffusion1}
\end{equation}
$E$ is the forced elevation at the given point, it is directly created from the elevation constraints of the feature curves.
$G$ describes the gradient or the slope and $L$ is a Laplace smoothing term.
The values $\alpha$ and $\beta$, with $0\leq\alpha\leq 1, 0\leq\beta\leq 1, \alpha+\beta\leq 1$, specify the weighting of the single term. On the plateaus of the feature curves we set $\alpha=0.9$ and $\beta=0$. This forces the elevation to match the desired height while adding a little bit of smoothing to it. During slopes of the feature curve, we use $\alpha=0$ and $\beta=0.5$. By that, both the slope constraints and smoothing constraints are satisfied. Outside of the influence of the feature curves, we set $\alpha=0, \beta=0$.

The gradient term is defined in the following way:
\begin{equation}
	G^{k+1}[i,j] = N^k[i,j]+G[i,j] \newline
\label{eq:Diffusion2}
\end{equation}
\begin{equation}
	N^k[i,j] = \textbf{n}_x^2 H^k[i-\text{sign}(\textbf{n}_x),j] + \textbf{n}_y^2 H^k[i, j-\text{sign}(\textbf{n}_y)]
\label{eq:Diffusion3}
\end{equation}
The gradient equation tend to satisfy the provided gradient $G$. $G$ is directly calculated from $\sin(\alpha_l)$ and $\sin(\alpha_r)$, i.e. from the slope angles defined at the control points.
The vector $\textbf{n}=(\textbf{n}_x,\textbf{n}_y)$ describes the normalized direction of the gradient. This vector is orthogonal to the feature curve when projected in the plane and point away from the curve. In Fig. \ref{fig:featsExist2}, some of them are displayed as gray lines going away from the plateau.

@@

This case happens quite often as seen in Fig. \ref{fig:diffusion3}. Here we define a new feature curve right next to an existing mountain. Although the curve's slope does not touch the existing terrain, the smoothing term inside the diffusion solver also increases the height outside of the feature curve. This leads to a growth of the mountain on the left.
From a designer's perspective, this behaviour might be not wanted because it is hard to predict and control.

\begin{figure*}
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{images/diffusionError1}
		\caption{original terrain with new feature curve}
		\label{fig:diffusionError1}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{images/diffusionError2}
		\caption{inference, mountain on the left side also grows}
		\label{fig:diffusionError2}
	\end{subfigure}
	\caption{Inferences between new feature curves and existing terrain features}\label{fig:diffusion3}
\end{figure*}

\section{Hydraulic Erosion}\label{HydraulicErosion}
TODO

>>>>

Each feature curve consists of two or more control points. Each control point has the following properties:
\begin{itemize}
	\item a position $(x,y,z)$ in the world
	\item a boolean if it constraints elevation
	\item a float $plateau$, specifying the size of the flat plateau on top of the curve
	\item four floats $s_l, \varphi_l, s_r, \varphi_r$ specifying the size and angle of the slopes left and right of the curve
\end{itemize}
In the example in Fig. \ref{fig:diffusion1}, the slope sizes are zero, so no slope constraints are applied.

@@

\subsection{Diffusion solver}\label{DiffusionSolver}
The task of the diffusion solver is to modify the terrain so that it remains smooth while preserving the feature constraints.
The diffusion solver iteratively updates the terrain using the following recursion: Let $H^i$ be the height map at the i-th iteration.
\begin{equation}
\begin{array}{rcl}
	H^{k+1}[i,j] &=& \alpha[i,j] E[i,j]  \\
	             &+& \beta[i,j] G^{k+1}[i,j]  \\
							 &+& (1-\alpha[i,j]-\beta[i,j]) L^{k+1}[i,j]
\end{array}
\label{eq:Diffusion1}
\end{equation}
$E$ is the forced elevation at the given point, it is directly created from the elevation constraints of the feature curves.
$G$ describes the gradient or the slope and $L$ is a Laplace smoothing term.
The values $\alpha$ and $\beta$, with $0\leq\alpha\leq 1, 0\leq\beta\leq 1, \alpha+\beta\leq 1$, specify the weighting of the single term. On the plateaus of the feature curves we set $\alpha=0.9$ and $\beta=0$. This forces the elevation to match the desired height while adding a little bit of smoothing to it. During slopes of the feature curve, we use $\alpha=0$ and $\beta=0.5$. By that, both the slope constraints and smoothing constraints are satisfied. Outside of the influence of the feature curves, we set $\alpha=0, \beta=0$.

The gradient term is defined in the following way:
\begin{equation}
	G^{k+1}[i,j] = N^k[i,j]+G[i,j] \newline
\label{eq:Diffusion2}
\end{equation}
\begin{equation}
	N^k[i,j] = \textbf{n}_x^2 H^k[i-\text{sign}(\textbf{n}_x),j] + \textbf{n}_y^2 H^k[i, j-\text{sign}(\textbf{n}_y)]
\label{eq:Diffusion3}
\end{equation}
The gradient equation tend to satisfy the provided gradient $G$. $G$ is directly calculated from $\sin(\varphi_l)$ and $\sin(\varphi_r)$, i.e. from the slope angles defined at the control points.
The vector $\textbf{n}=(\textbf{n}_x,\textbf{n}_y)$ describes the normalized direction of the gradient. This vector is orthogonal to the feature curve when projected in the plane and point away from the curve. In Fig. \ref{fig:featsExist2}, some of them are displayed as gray lines going away from the plateau.

@@

This case happens quite often as seen in Fig. \ref{fig:diffusion3}. Here we define a new feature curve right next to an existing mountain. Although the curve's slope does not touch the existing terrain, the smoothing term inside the diffusion solver also increases the height outside of the feature curve. This leads to a growth of the mountain on the left (the tip of the mountain is now much higher than the control point, before it was about the same height).
From a designer's perspective, this behavior might be not wanted because it is hard to predict and control.

\begin{figure*}
	\centering
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/diffusionError1}
		\caption{original terrain with new feature curve}
		\label{fig:diffusionError1}
	\end{subfigure}
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/diffusionError2}
		\caption{inference, left mountain is modified}
		\label{fig:diffusionError2}
	\end{subfigure}
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/diffusionErrorSolved}
		\caption{limiting smoothing to remove inference}
		\label{fig:diffusionError3}
	\end{subfigure}
	\caption{Inferences between new feature curves and existing terrain features}\label{fig:diffusion3}
\end{figure*}

\subsection{Extension of the solver}
We propose now another solution that both keeps existing terrain features while it preserves the local control and removes the inferences mentioned before.
The idea is to limit the smoothing term $L$ to influence only a customizable region around a feature curve.
For this, we add two new parameters, $l_l$ and $l_r$, to each control point. They specify the local radius of an envelope around the plateau and slope of the feature curve.
It not only extends along the slope, but also over the start and end of the feature curve.

We then set $\gamma$ to 1 inside this envelope and to 0 outside and modify equation \ref{eq:Diffusion1} to include $\gamma$:
\begin{equation}
\begin{array}{rcl}
	H^{k+1}[i,j] &=& \alpha[i,j] E[i,j]  \\
	             &+& \beta[i,j] G^{k+1}[i,j]  \\
							 &+& \gamma[i,j] (1-\alpha[i,j]-\beta[i,j]) L^{k+1}[i,j]
\end{array}
\label{eq:Diffusion1_mod}
\end{equation}

Because we no limit the smoothing area, we can apply the solver directly on the original height map without smoothing out existing features. There is no need anymore to solve the diffusion equation on a separate map with relative heights as done in the previous section.

In Fig. \ref{fig:diffusionError3} you see the problematic situation again, but this time, the smoothing region is bounded as described before. The border of the smoothing envelope is visualized by the black lines. Note that the old mountain is preserved and the new feature is blended into the terrain without discontinuities.

\section{Hydraulic Erosion}\label{HydraulicErosion}
Now the shape of the terrain is fully defined, and the next step is to increase the realism by performing erosion.
